---
output:
  github_document:
    toc: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
```


# cQL

This R package `cQL` implements a novel clustered Q-learning method with the $M$-out-of-$N$ cluster bootstrap for making inference on tailoring variables in a two-stage clustered SMART (cSMART).

The package is designed for analyzing one completed cSMART dataset at a time. Given a user-supplied dataset and user-specified stage-1 and stage-2 Q-functions, the main function `cQL()` could:

1. fits the stage-2 Q-function;
2. estimates the degree of non-regularity;
3. selects the stage-1 resample size `M`;
4. constructs the stage-1 pseudo-outcome; and
5. returns bootstrap-based inference for both stages.

The output includes:

- regression coefficient estimates;
- confidence intervals at the requested level (95% by default);
- bootstrap standard errors;
- bootstrap p-values;
- significance stars; and
- a stage summary showing `N`, `N_rand`, and `M` for each stage.

## Installation

To install this package from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("SelinaSong0412/cQL")
```

For local development:

```r
devtools::install(".")
```

Then load the package:

```r
library(cQL)
```

## Data format required by `cQL()`

Before using the package, prepare the cSMART data in the following format.

1. The dataset should be at the individual level, so each row is one individual.
2. The dataset should contain a cluster id column.
3. The dataset should contain one stage-1 treatment column and one stage-2
   treatment column.
4. The dataset should contain the final observed outcome column `Y`.
5. If the dataset also contains an observed stage-1 intermediate outcome, that
   column should be named `Y1`.
6. If the cSMART is a Design II or III trial with limited second-stage
   re-randomization, then the stage-2 treatment should be `NA` for clusters that
   were not re-randomized at stage 2.
7. Cluster-level variables such as the treatments and candidate tailoring
   variables should be repeated across all individuals within the same cluster.

In other words:

- for Design I cSMART data, the stage-2 treatment column is fully observed;
- for Design II or III cSMART data, the stage-2 treatment column contains
  `NA` for non-re-randomized clusters.
- if the user only has the final observed outcome `Y`, then the stage-1
  pseudo-outcome is built from the stage-2 pseudo-outcome alone;
- if the user has both `Y1` and `Y`, then the stage-1 pseudo-outcome is built
  as `Y1 +` the stage-2 pseudo-outcome.

The package can work with any binary coding of the two treatment columns. If
the treatments are not already coded as `-1` and `1`, `cQL()` internally
recodes them and stores the mapping in the fitted object.

The stage-1 formula should always use `Y1_tilde` as its response, for example
`Y1_tilde ~ X1 * A1`. If `Y1` is available, the stage-2 formula may include
`Y1` as a predictor.

The function also lets the user choose the significance level through `alpha`.
If `alpha = NULL`, the package uses the default `alpha = 0.05`, which gives
95% confidence intervals. For example, set `alpha = 0.10` for 90% confidence
intervals.

## Recommended workflow

The recommended workflow for your own data is:

1. prepare the data in the required format;
2. specify the stage-2 and stage-1 Q-functions;
3. fit `cQL()`;
4. inspect `fit$stage_summary` to see how many clusters are used at each stage
   and what `M` was selected; and
5. inspect `fit$stage2` and `fit$stage1` to interpret the stage-specific
   tailoring effects.

## Example 1: Design I cSMART with full second-stage re-randomization

Below, we generate a toy Design I cSMART dataset. In this design, all clusters
are re-randomized at stage 2, so the stage-2 treatment column has no missing
values.

### Step 1: generate a full re-randomization dataset

```{r}
design1_data <- simulate_csmart_data(
  n_clusters = 40,
  cluster_size = 20,
  rerandomization = "full",
  seed = 111
)

head(design1_data)
table(is.na(design1_data$A2))
```

The table above should show that `A2` is never missing, which is what we
expect for Design I.

### Step 2: specify the Q-functions

```{r}
s2_formula <- Y ~ X1 * A1 + A1 * A2 + A2:X2
s1_formula <- Y1_tilde ~ X1 * A1
```

In this example:

- the stage-2 candidate tailoring variables are `A1` and `X2`, because they
  interact with `A2`;
- the stage-1 candidate tailoring variable is `X1`, because it interacts with
  `A1`.

### Step 3: fit the clustered Q-learning analysis

```{r}
fit_design1 <- cQL(
  data = design1_data,
  stage2_formula = s2_formula,
  stage1_formula = s1_formula,
  cluster = "cluster_id",
  stage1_treat = "A1",
  stage2_treat = "A2",
  stage2_tailoring_vars = c("A1", "X2"),
  working_correlation = "exchangeable",
  alpha = NULL,
  n_boot = 150,
  seed = 412,
  verbose = FALSE
)
```

Here we use the exchangeable working correlation model by specifying
`working_correlation = "exchangeable"`, which is also the default. If desired,
the user can instead set `working_correlation = "independence"`.

### Step 4: inspect how the algorithm works internally

```{r}
fit_design1$stage_summary
```

How to read this output:

- `N` is the total number of clusters in the dataset.
- `N_rand` is the number of clusters randomized at that stage.
- `M` is the bootstrap resample size used at that stage.

For Design I:

- at stage 2, all clusters are randomized, so `N_rand = N`, and stage 2 uses a
  cluster bootstrap over those randomized clusters, so `M = N_rand`;
- at stage 1, all clusters are also randomized, but the manuscript's
  M-out-of-N rule may choose `M < N` when non-regularity is present.

### Step 5: inspect the stage-2 inference

```{r}
fit_design1$stage2
```

To interpret the stage-2 table, focus especially on the terms involving `A2`.
If an interaction involving `A2` has a confidence interval that excludes zero
and a small p-value, that suggests the corresponding variable may be useful as
a stage-2 tailoring variable.

### Step 6: inspect the stage-1 inference

```{r}
fit_design1$stage1
```

The stage-1 table is built after constructing the stage-1 pseudo-outcome from
the fitted stage-2 model. In this example there is no observed `Y1`, so
`Y1_tilde` is the stage-2 pseudo-outcome itself. Terms involving `A1` are the
stage-1 tailoring effects of interest.

## Example 2: Design II or III cSMART with limited second-stage re-randomization

Now we generate a toy cSMART dataset in which only a subset of clusters is
re-randomized at stage 2. This mimics the data format required for Design II
or III cSMARTs.

### Step 1: generate a partial re-randomization dataset

```{r}
design23_data <- simulate_csmart_data(
  n_clusters = 40,
  cluster_size = 20,
  rerandomization = "nonresponder",
  p_rerand = 0.7,
  seed = 222
)

head(design23_data)
```

Here the stage-2 treatment column contains `NA` for clusters that were not
re-randomized at stage 2. This is the key formatting rule for Design II or III
data.

The package does not need separate arguments telling it whether the
re-randomized clusters were responders or non-responders. The crucial input is
simply:

- one row per individual; and
- `NA` in the stage-2 treatment column for clusters not re-randomized.

### Step 2: use the same Q-functions

```{r}
s2_formula
s1_formula
```

### Step 3: fit the clustered Q-learning analysis

```{r}
fit_design23 <- cQL(
  data = design23_data,
  stage2_formula = s2_formula,
  stage1_formula = s1_formula,
  cluster = "cluster_id",
  stage1_treat = "A1",
  stage2_treat = "A2",
  stage2_tailoring_vars = c("A1", "X2"),
  working_correlation = "exchangeable",
  alpha = NULL,
  n_boot = 150,
  seed = 412,
  verbose = FALSE
)
```

### Step 4: inspect the stage summary

```{r}
fit_design23$stage_summary
```

This output is often the easiest way to understand what the algorithm is doing
internally.

For a partial re-randomization design:

- the stage-2 row should have `N_rand < N`, because only some clusters are
  randomized at stage 2;
- the stage-2 bootstrap resamples those stage-2 randomized clusters, so the
  stage-2 row has `M = N_rand`;
- the stage-1 row reports the manuscript-selected stage-1 `M` for the
  M-out-of-N bootstrap.

### Step 5: inspect the stage-2 inference

```{r}
fit_design23$stage2
```

The interpretation is the same as before, but now the stage-2 regression is
fit only to the clusters that were actually re-randomized at stage 2.

### Step 6: inspect the stage-1 inference

```{r}
fit_design23$stage1
```

For limited second-stage re-randomization, the manuscript's stage-1
pseudo-outcome rule is used:

- for clusters re-randomized at stage 2, the pseudo-outcome uses the fitted
  stage-2 Q-function;
- for clusters not re-randomized at stage 2, the pseudo-outcome equals the
  observed outcome.

## Example 3: Design II cSMART with an observed stage-1 outcome `Y1`

This example uses limited second-stage re-randomization again, but now the
input data include both an observed stage-1 intermediate outcome `Y1` and the
final outcome `Y`.

### Step 1: create a Design II-style dataset with both `Y1` and `Y`

```{r}
design23_y1_data <- simulate_csmart_data(
  n_clusters = 40,
  cluster_size = 20,
  rerandomization = "nonresponder",
  p_rerand = 0.7,
  seed = 333
)

set.seed(412)
cluster_effect_y1 <- stats::rnorm(
  length(unique(design23_y1_data$cluster_id)),
  sd = 0.25
)
names(cluster_effect_y1) <- as.character(unique(design23_y1_data$cluster_id))

design23_y1_data$Y1 <- with(
  design23_y1_data,
  0.4 +
    0.5 * X1 +
    0.3 * A1 +
    0.2 * X1 * A1 +
    cluster_effect_y1[as.character(cluster_id)] +
    stats::rnorm(nrow(design23_y1_data), sd = 0.6)
)

design23_y1_data$Y <- design23_y1_data$Y + 0.35 * design23_y1_data$Y1

head(design23_y1_data[c("cluster_id", "patient_id", "X1", "A1", "X2", "A2", "Y1", "Y")])
```

### Step 2: specify the Q-functions

```{r}
s2_formula_y1 <- Y ~ Y1 + X1 * A1 + A1 * A2 + A2:X2
s1_formula_y1 <- Y1_tilde ~ X1 * A1
```

Here `Y1` is allowed in the stage-2 formula, but the stage-1 formula still uses
`Y1_tilde` as its response.

### Step 3: fit the clustered Q-learning analysis

```{r}
fit_design23_y1 <- cQL(
  data = design23_y1_data,
  stage2_formula = s2_formula_y1,
  stage1_formula = s1_formula_y1,
  cluster = "cluster_id",
  stage1_treat = "A1",
  stage2_treat = "A2",
  stage2_tailoring_vars = c("A1", "X2"),
  working_correlation = "exchangeable",
  alpha = NULL,
  n_boot = 150,
  seed = 412,
  verbose = FALSE
)
```

### Step 4: inspect the results

```{r}
fit_design23_y1$stage_summary
fit_design23_y1$stage2
fit_design23_y1$stage1
```

In this scenario the stage-1 pseudo-outcome is:

- `Y1 +` the fitted stage-2 pseudo-outcome for clusters re-randomized at stage 2;
- `Y1 + Y` for clusters not re-randomized at stage 2.

## Applying `cQL()` to your own data

After your data are prepared, the analysis for your own cSMART dataset will
look like this:

```r
my_fit <- cQL(
  data = my_csmart_data,
  stage2_formula = Y ~ X1 * A1 + A1 * A2 + A2:X2,
  stage1_formula = Y1_tilde ~ X1 * A1,
  cluster = "cluster_id",
  stage1_treat = "A1",
  stage2_treat = "A2",
  stage2_tailoring_vars = c("A1", "X2"),
  working_correlation = "exchangeable",
  alpha = NULL,
  n_boot = 1000,
  fixed_xi = 0.025
)

my_fit$stage_summary
my_fit$stage2
my_fit$stage1
```

In practice:

- start with `my_fit$stage_summary` to see the cluster counts used by the
  algorithm;
- then inspect `my_fit$stage2` to evaluate candidate stage-2 tailoring
  variables;
- then inspect `my_fit$stage1` to evaluate candidate stage-1 tailoring
  variables.

If your data contain an observed stage-1 outcome `Y1`, keep the stage-1 formula
as `Y1_tilde ~ ...` and optionally include `Y1` in `stage2_formula`.

If you want a confidence level other than 95%, specify `alpha` directly. For
example, use `alpha = 0.10` to request 90% confidence intervals.

## Notes

- The current package targets the two-stage cSMART setting developed in the
  manuscript.
- The stage-2 formula must include the main effect of the stage-2 treatment.
- Every interaction involving the stage-2 treatment must be a two-way
  interaction with one of the variables listed in `stage2_tailoring_vars`.
- In the printed stage summary, the stage-2 row has `M = N_rand` because stage
  2 uses the full cluster bootstrap over the stage-2 randomized clusters, whereas
  the stage-1 row uses the selected M-out-of-N resample size. See more technical detail about the algorithm design in [the original manuscript](https://arxiv.org/abs/2505.00822)
